---
title: "Assignment 5: Data Visualization"
author: "Jazmine Pritchett"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1.  Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 

#project set up by install of packages and importing of data.

library(tidyverse)
library(lubridate)
library (cowplot)
library(here)
getwd()
      
PeterPaul.NTL <- read.csv(
  file=here("Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"), 
  stringsAsFactors = TRUE)

NR.Litter <- read.csv(file=here("Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"),
                    stringsAsFactors = TRUE)


#2 

#Dates were read as factors.

class(PeterPaul.NTL$sampledate)

PeterPaul.NTL$sampledate <- ymd(PeterPaul.NTL$sampledate)

class(NR.Litter$collectDate)

NR.Litter$collectDate <- ymd(NR.Litter$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3

#Set a theme that aesthetically pleasing and that doesn't take away 
#from data values.

library(ggthemes)

my_theme <- theme_base() + theme (legend.background = element_rect(
      color='black',
      fill = 'darkgrey'),
    axis.ticks = element_line(
      color='darkgrey', linewidth = 0.5))

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 

#Creation of graph with edits to title for aesthetics. 

NTLplot <- ggplot(PeterPaul.NTL, aes(x = po4, y = tp_ug)) + 
my_theme + geom_point(aes(color = lakename), na.rm = TRUE) +  
xlim(0,25) + ylim(0, 125)+
geom_smooth(color = 'black', method=lm) + labs(x= "Phosphate
(ug)", y= "Total Phosphorus (ug)", title = "Total Phosphorus by Phosphate")
print(NTLplot)

theme_set(my_theme)

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tip:
* Recall the discussion on factors in the previous section as it may be helpful here. 
* R has a built-in variable called `month.abb` that returns a list of months;see <https://r-lang.com/month-abb-in-r-with-example>

```{r Create boxplots}
#5 

TempPlot <- PeterPaul.NTL %>% ggplot(aes(y = temperature_C, x = factor(month, levels = 1:12, 
labels=month.abb ))) + 
scale_x_discrete(name="Month", drop =FALSE)+ geom_boxplot(aes(color = lakename)) + labs(x= "Month", 
y= "Temperature (C)")
print(TempPlot)

TPPlot <- PeterPaul.NTL %>% ggplot(aes(y = tp_ug, x = factor(month, levels = 1:12, 
labels= month.abb
))) + scale_x_discrete (name="Month", drop =FALSE)+ geom_boxplot(aes(color = lakename))+ 
labs(x= "Month", y= "Phosphorus (ug)")
print(TPPlot)

TNPlot <-PeterPaul.NTL %>% ggplot(aes(y = tn_ug, x = factor(month, levels = 1:12, labels=month.abb 
))) + scale_x_discrete(name="Month", drop =FALSE)+ geom_boxplot(aes(color = lakename))+ 
labs(x= "Month", y= "Nitrogen (ug)")
print(TNPlot)


plot_grid(TempPlot + theme(legend.position = "top"), #legend included at the top for aligned axis.
TPPlot + theme(legend.position="none"), TNPlot + theme(legend.position="none"),
nrow=3, align= 'v', #align = v aligned
#vertical axis plot when placing in the legend at the top.
rel_heights = c(1.25,1))


```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: Each lake is more prominent over warmer/summer months (May-Septemeber). There is also not much variance for each lake across variables and the months.

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6

Needles_Group <- ggplot(subset(NR.Litter, functionalGroup == "Needles"), aes(x = collectDate,
y = dryMass))+ geom_point(aes(color = nlcdClass))+ labs(x= "Collection Date", y= "Dry Mass (g)")
print(Needles_Group)


#7

Needles_Facet <- ggplot(subset(NR.Litter, functionalGroup == "Needles"), aes(x = collectDate,
y = dryMass))+ geom_point() + facet_wrap(vars(functionalGroup), nrow = 3) + labs(x= "Collection Date", 
y= "Dry 
Mass (g)") + facet_grid( functionalGroup ~ nlcdClass) #plotting Needles by facets of nlcd classes.
print(Needles_Facet)

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer:While plot 6 is in color, values are overlapping. Because of this, I think plot 7 that includes facets helps the viewer easily compare each class of data.