---
title: "Assignment 3: Data Exploration"
author: "Jazmine Pritchett"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Check your working directory, load necessary packages (tidyverse, lubridate), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r Setting up data, message=FALSE}
getwd()
#My working directory is in my class folder where my R project is located.
#I do not need to change it, but if I did I coulde use an absolute file path or go to the
#engine to set one.

#tidyverse and lubridate are already installed, so I will simply load them

library(tidyverse)
library(lubridate)

#Imported data and assigned names to the dataset. Assigned strings as 
#factors to true to convert character/categorical values. 

Neonics <-read.csv("./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv", stringsAsFactors = T)

Litter <- read.csv("./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv",stringsAsFactors = T)

#Checks to ensure data looks correct and that column header 
#names are not taking the first line.

colnames(Neonics)
colnames(Litter)
```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer:There are environmental impacts of neonicotinoids that disrupt food webs. Some effects include impacts on reproduction and impairements to function for pollinating insects. 

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer:Litter and woody debris is important for carbon cycling in forests. These fallen trees that decay help restore nutrients to the forest environment and help it thrive. There is also importance of the role in providing habitat. 

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1. Sampling occurs at sites that have woody vegetation >2m tall.
 2. There is one litter trap pair for every 400 m^2 plot area meaning overall there is 1-4 trap pairs per plot. 
 3. Tower plot locations are selected randomly within 90% flux footprint of primary and secondary airsheds. 

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r Dimensions of Neonics Data}
#Dimension tells me I have 4623 rows and 30 columns in the
#Neonics dataset. 

dim(Neonics)
```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?

```{r Common Neonic Effects}
#By taking summary of the specific column 'Effect' able to see 
#how many times each effect is by the summarized data. 

summary(Neonics$Effect)
```

> Answer: Most common effects studied include population, mortality, behavior, feeding behavior, and development. These are probably particularly of interest because as state before neonicotinoids affect function and reproductability. 

7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: The `sort()` command can sort the output of the summary command...]

```{r Common Neonic Species}
sort(summary(Neonics$Species.Common.Name))
```

> Answer: 1. Honey Bee, 2. Parasitic Wasp, 3. Buff Tailed Bumblebee, 4. Carniolan Honey Bee, 5. Bumble Bee, 6. Italian Honey Bee. These top commonly studied species are flying pollinators, most being bees. These might be of most interest because as said before neonicotinoids most commonly affect pollinaators when they come in contact with plants.

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric?

```{r Concentration Class Check}
class(Neonics$Conc.1..Author.)
```

> Answer: It is a factor. This column tells how the concentration was prepared or what is included, such as an active ingredient or formulation.  

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r Studies by Publication Year Plot}
ggplot(Neonics)+geom_freqpoly(aes(x= Publication.Year), bins = 50)

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r Test Location Plot}
ggplot(Neonics)+geom_freqpoly(aes(x= Publication.Year, color = Test.Location), bins=50) +
  theme(legend.position = "top")

#For aesthetic purposes, the legend was moved to the top. Each record for test location has 
#been assigned a color.  
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: Most common is the lab followed by natural field. Overtime both of these locations have increased in count, specially for the lab. For other variables, they have remained constant.

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r Endpoint Bar Graph}
ggplot(Neonics, aes(x=Endpoint)) + geom_bar() + theme(axis.text.x = element_text(angle = 90, 
vjust = 0.5, hjust=1))
```

> Answer: The two most common endpoints are LOEL and NOEL. LOEL and NOEL are the database usage of terrestial. LOEL is the lowest-observable-effect-level where  lowest dose (concentration) producing effects that were significantly different from responses of controls. NOEL is no-observable-effect-level where highest dose (concentration) producing effects not significantly different from responses of controls according to author's reported statistical test. 

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r collectDate Class}

class(Litter$collectDate)

Litter$collectDate <- c('2018-08-02','2018-08-30')

Litter$collectDate <- as.Date(Litter$collectDate, format = '%Y-%m-%d')

unique(Litter$collectDate, 2018-08)

#The class was a factor, not a date. Convert collect date to format year (four digit), month, day.

```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r Unique versus Summary}
unique(Litter)
summary(Litter)
```

> Answer: 188 plots were sampled when looking at the data in the environment. 'unique' gets rid of duplicate counts in data to help get a clearer understanding of the data, there weren't any so there was no difference. 

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r functionalGroup Bar Graph}
ggplot(Litter, aes(x=functionalGroup)) + geom_bar() + theme(axis.text.x = element_text(angle = 90, 
vjust = 0.5, hjust=1))
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r dryMass by functionalGroup Plots}
ggplot(Litter) + geom_boxplot(aes(x = functionalGroup, y = dryMass))

ggplot(Litter) + geom_violin(aes(x = functionalGroup, y = dryMass))
```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: The box plot is more effective at visualizing the data compared to the violin plot as it effectively shows the distribution of the range of the data vertically. The violin plot widths aren't very visual horizontally. 

What type(s) of litter tend to have the highest biomass at these sites?

> Answer: Needles and somewhat mixed and twigs/branches. 
